Open archaeology, open source?

### 2024-01-30 analysis/_03-data_methodology.qmd
Mostly minor but necessary fixes to code and text
Provided a little explanation for the manual licenses file, included a missing value to the licenses file, corrected some citations, added section slugs and encoded links to sections, removed various completed or unnecessary todos and notes, removed dendrogram, verified and updated characterization of clusters, fixed references to graph objects

# Data and methodology

@@

We present an exploratory analysis of open-archaeo [[open-archaeo.info](open-archaeo.info), @batist2023], a directory of `r nrow(oarch)` <!-- TODO: make this consistent with what appears in Section 4, based on the filter --> pieces of open source archaeological software and other digital resources.

@@

The open-archaeo dataset and the supplemental data from the GitHub API provide us with a rich resource to explore the makeup of the archaeological software development scene.
We apply exploratory data analysis methods to break down and visualize the data, so that we may ascertain overall trends [@tukey1977].
In Section 4, <!-- encode section, see https://quarto.org/docs/authoring/cross-references.html#sections --> our focus is on examining the general state of open source archaeological software and resource development.
In Section 5, <!-- encode section --> we refine our analysis to examine development processes, with specific focus on collaborative experiences.
Finally, in Section 6, <!-- encode section --> we apply network analysis methods to investigate the formation of broader collaborative communities.
Our analyses combine to support our objectives of understanding what kinds of software and resources archaeologists are making, and how they create these tools in response to specific needs and use-cases, as afforded by offline social and professional connections, and within the context of an emerging community of practice.

The quantitative analyses and figures presented here were generated with `r version$version.string` (@Manual).
The full data and code is available in the compendium that accompanies this paper (@Compendium).

>>>>

# Data and methodology {#sec-data-methodology}

@@

We present an exploratory analysis of open-archaeo [[open-archaeo.info](open-archaeo.info), @batist2023], a directory of `r nrow(oarch)` pieces of open source archaeological software and other digital resources.

@@

The open-archaeo dataset and the supplemental data from the GitHub API provide us with a rich resource to explore the makeup of the archaeological software development scene.
We apply exploratory data analysis methods to break down and visualize the data, so that we may ascertain overall trends [@tukey1977].
In @sec-open-archaeology, our focus is on examining the general state of open source archaeological software and resource development.
In @sec-collaboration, we refine our analysis to examine development processes, with specific focus on collaborative experiences.
Finally, in @sec-network, we apply network analysis methods to investigate the formation of broader collaborative communities.
Our analyses combine to support our objectives of understanding what kinds of software and resources archaeologists are making, and how they create these tools in response to specific needs and use-cases, as afforded by offline social and professional connections, and within the context of an emerging community of practice.

The quantitative analyses and figures presented here were generated with `r version$version.string` [@rcoreteam2023].
The full data and code is available in the compendium that accompanies this paper [@roe2023] <!-- reference should be updated to reflect the actual version number and date -->.

### 2024-01-30 analysis/_04-open_archaeology.qmd
Mostly minor but necessary fixes to code and text
Provided a little explanation for the manual licenses file, included a missing value to the licenses file, corrected some citations, added section slugs and encoded links to sections, removed various completed or unnecessary todos and notes, removed dendrogram, verified and updated characterization of clusters, fixed references to graph objects

# Open archaeology

@@

```{r tbl-categories}
#| tbl-cap: Categories of open archaeology projects included in open-archaeo.
# TODO: Include sum for the "Software" and "Documents" supercategories, and make them appear in bold.
tribble(
  ~category,                               ~kind,       ~scope,
  "Packages and libraries",                "Software", "Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.",
  "Standalone software",                   "Software", "Software that may be operated without needing to first access an underlying platform.",
  "Scripts",                               "Software", "Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.",
  "Specifications, protocols and schemas", "Documents", "A formal data structure or framework intended to be used as a model.",
  "Products",                              "Documents", "Stable outcomes of creative work.",
  "Guides",                                "Documents", "An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.",
  "Lists and datasets",                    "Documents", "A series of consistently organized observations assembled with purpose."
) |>
  left_join(oarch, by = "category") |>
  count(category, kind, scope) |>
  group_by(kind) |>
  arrange(-n) |>
  rename_with(tools::toTitleCase) |>
  tt()
```

@@

```{r data-licenses}
oarch_licenses <- read_csv(here("analysis/data/raw_data/open-archaeo_licenses.csv"))
oarch_licenses |>
  select(gh_repo = url, other_license = license) |>
  right_join(oarch, by = "gh_repo") |>
  hoist(gh_repo_info, license = list("license", 1, "spdx_id")) |>
  mutate(license = if_else(license == "NOASSERTION", other_license, license)) |>
  select(-other_license) ->
  oarch

@@

```{r tbl-repo-licenses}
#| tbl-cap: Summary of licenses used by projects with GitHub repositories
# TODO: group small categories into "other"
oarch |>
  mutate(license = str_remove(license, "-?\\d\\.\\d$")) |>
  count(license) |>
  mutate(license = replace_na(license, "None detected")) |>
  arrange(-n) |>
  rename_with(tools::toTitleCase) |>
  tt()
```

>>>>

# Open archaeology {#sec-open-archaeology}

@@

```{r tbl-categories}
#| tbl-cap: Categories of open archaeology projects included in open-archaeo.
tribble(
  ~category,                               ~kind,       ~scope,
  "Packages and libraries",                "Software", "Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.",
  "Standalone software",                   "Software", "Software that may be operated without needing to first access an underlying platform.",
  "Scripts",                               "Software", "Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.",
  "Specifications, protocols and schemas", "Documents", "A formal data structure or framework intended to be used as a model.",
  "Products",                              "Documents", "Stable outcomes of creative work.",
  "Guides",                                "Documents", "An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.",
  "Lists and datasets",                    "Documents", "A series of consistently organized observations assembled with purpose."
) |>
  left_join(oarch, by = "category") |>
  count(category, kind, scope) |>
  group_by(kind) |>
  arrange(-n) |>
  rename_with(tools::toTitleCase) |>
  tt()
```

@@

```{r data-licenses}
# open-archaeo_licenses.csv contains a list of licenses that had to be manually verified. This is necessary due to the fact that GitHub and CRAN have different conventions for naming the files containing the license.
oarch_licenses <- read_csv(here("analysis/data/raw_data/open-archaeo_licenses.csv"))
oarch_licenses |>
  select(gh_repo = url, other_license = license) |>
  right_join(oarch, by = "gh_repo") |>
  hoist(gh_repo_info, license = list("license", 1, "spdx_id")) |>
  mutate(license = if_else(license == "NOASSERTION", other_license, license)) |>
  select(-other_license) ->
  oarch

@@

```{r tbl-repo-licenses}
#| tbl-cap: Summary of licenses used by projects with GitHub repositories
oarch |>
  mutate(license = str_remove(license, "-?\\d\\.\\d$")) |>
  count(license) |>
  mutate(license = replace_na(license, "None detected")) |>
  arrange(-n) |>
  rename_with(tools::toTitleCase) |>
  tt()
```

### 2024-01-30 analysis/_06-network.qmd
Mostly minor but necessary fixes to code and text
Provided a little explanation for the manual licenses file, included a missing value to the licenses file, corrected some citations, added section slugs and encoded links to sections, removed various completed or unnecessary todos and notes, removed dendrogram, verified and updated characterization of clusters, fixed references to graph objects

# An emerging community of practice?

@@

```{r fig-graph-repo-user, eval=FALSE}
#| fig-cap: Interactive repository–user graph.
# TODO: is there a khroma scale that can accomodate this many groups?
g_repo_repo <- oarch_graph_repo_user |>
  ggraph(layout = "graphopt", mass = 100) +
  geom_edge_link(aes(group = contrib, alpha = n_contrib)) +
  geom_point_interactive(aes(x, y, colour = node_type, tooltip = node)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common contributors") +
  theme(panel.background = element_rect(fill = "white"))

@@

```{r fig-graph-repo-repo}
#| fig-cap: Interactive repository–repository graph. Repositories are connected by contributions by common users, with darker edges indicating more common contributors. Node colour indicates cluster membership according to .
# TODO: is there a khroma scale that can accomodate this many groups?
g_repo_repo <- oarch_graph_repo_repo |>
  ggraph(layout = "graphopt", mass = 100) +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, colour = group, tooltip = repo)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common contributors") +
  theme(panel.background = element_rect(fill = "white"))

```{r fig-graph-user-user, eval=FALSE}
#| fig-cap: Interactive user–user graph. Users are connected by contributions by common repositories, with darker edges indicating more common repositories. Node colour indicates cluster membership according to the edge–betweenness method [@Girvan Newman2002].
# TODO: is there a khroma scale that can accomodate this many groups?
g_repo_repo <- oarch_graph_user_user |>
  filter(degree > 170) |>
  ggraph(layout = "graphopt", mass = 100) +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, tooltip = user)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common contributors") +
  theme(panel.background = element_rect(fill = "white"))

<!-- Is there a network? How big is it? -->
Our data shows that there is a significant network of archaeologists collaborating on GitHub.
`r percent(graph_n_repos / graph_n_repos_total)` of repositories and `r percent(graph_n_users / graph_n_users_total)` of users in our dataset are connected to at least one other repository/user.
Of these, `r percent(max(subgraphs_n_repos$n) / graph_n_repos)` of repositories and `r percent(max(subgraphs_n_users$n) / graph_n_users)` of users belong to a single connected subgraph (@fig-graph-repo-repo).

```{r fig-graph-repo-repo-edge-betweenness-dendrogram}
# ZB Note: What we really want out of this is the identification of groups or clusters, as well as the relationships between clusters. In theory, the dendrogram provides the best visualization for this information. I wonder if we can improve it by colour-coding the algorithmically-identified clusters (which should appear side by side), and making the names of the repositories appear vertically aligned perpendicular to the x axis. Additionally, we can colour-code the names of the repositories with their corresponding colours, or indicate that they are part of a cluster using a bracket (like "{" but turned 180 degrees).
# @ZB: do we need both this and the graph viz. with clusters, though?
xx <- igraph::edge.betweenness.community(oarch_graph_repo_repo)
xx <- as.dendrogram(xx)
plot(xx)

# TODO: not working
#ggraph(xx, 'dendrogram', height = height)
```

<!-- Clustering within the network -->
To explore the topography of this network, we applied the edge–betweenness community detection method [@GirvanNewman2002] to delimit `r nrow(filter(clusters_n_repos, n > 1))` distinct clusters in the repository–repository network (@fig-graph-repo-repo-edge-betweenness-dendrogram).
While many of these clusters are interconnected, some discrete components containing between 2-7 repositories appear as distinct from a primary core.
The core clusters are characterized by repositories whose contributors commit to projects other than their own. 
The other peripheral clusters largely correspond with the work of single individuals, and sometimes also their close colleagues.
Peripheral clusters that are connected to core clusters by only a few relationships represent the sole (or perhaps initial) integration of lone developers into a broader community.

<!-- @ZB: Can you check that this is still valid, after we removed a lot of the genetics entries? -->
The three core clusters, have distinct characters.
One has a focus on archaeogenetics, which consists of a very well established collaborative network and general reliance on data modelling and processing tools. 
A second cluster is mostly centred on fieldwork-oriented data collection tools, and particularly the emergence of well-funded and well-supported dominant platforms that attract more attention than other independent projects scattered across the network. 
The third and most significant cluster includes a smorgasbord of projects whose contributors share varied interests. 
The emphasis in this latter cluster is on the formation of a central software development community rather than on any specific topic of work. 
Many of the projects represented in this third cluster emerge from underlying professional partnerships, namely research labs (e.g. ISAA-Kiel) and special interest groups (SSLA).

@@

The highest mean centrality is seen in repositories that contain lists/datasets, standalone software, or packages/libraries, and in repositories based on Python, R or QGIS [@fig-repo-betweenness-category-platform].
Interestingly, these trends depart from the observed popularity of different categories/platform in the open-archaeo dataset as a whole (see section XXX<!-- TODO: How do you do section references in Quarto? -->): standalone software is more central than packages/libraries, even though there are more of the latter by a significant margin.
Similarly, Python packages tend to be more central than R packages, despite being the minority language.
<!-- @ZB: I think this is interesting, but I couldn't really think of a general observation to make here. Maybe you can? -->

@@

<!-- Predictors of user centrality -->
Centrality to the user–user graph is weakly predicted by a user's overall rate of activity, as measured by their total number of contributions [@fig-user-betweenness-contribs]. 
<!-- @ZB: Have softened the following, not because I think it's wrong, just because I'm a bit uncomfortable making strong claims without a reproducible analysis to back it up. 
Also N.B.: I don't think they write code for a living. I don't know of anybody that rights code for a living in archaeology. I think it's the opposite, actually. -->
We did not collect demographic data on users that appear in our dataset, but based on our own knowledge of the community we can observe that those highly central to the network tend to be employed in (junior) academic positions, or in a few cases in cultural heritage authorities, rather than specifically as research software engineers.
Such positions tend to not to actively reward or encourage software development, at least not on a par with more traditional academic outputs, and are increasingly precarious.
This obviously poses a serious risk to the sustainability and growth of open source software in archaeology: if the people who occupy central positions in the network cease to be active, then it is likely that the overall network would fragment.
Assessing and mitigating this risk should be a high priority for future research in this area.

>>>>

# An emerging community of practice? {#sec-network}

@@

```{r fig-graph-repo-user}
#| fig-cap: Interactive repository–user graph.
# TODO: is there a khroma scale that can accomodate this many groups?
g_repo_user <- oarch_graph_repo_user |>
  ggraph(layout = "graphopt", mass = 100) +
  geom_edge_link(aes(group = contrib, alpha = n_contrib)) +
  geom_point_interactive(aes(x, y, colour = node_type, tooltip = node)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common contributors") +
  theme(panel.background = element_rect(fill = "white"))

girafe(ggobj = g_repo_user)
```

```{r fig-graph-repo-repo}
#| fig-cap: Interactive repository–repository graph. Repositories are connected by contributions by common users, with darker edges indicating more common contributors. Node colour indicates cluster membership according to the edge–betweenness method [@girvan2002]. Excludes isolate nodes.
# TODO: is there a khroma scale that can accomodate this many groups?
g_repo_repo <- oarch_graph_repo_repo |>
  ggraph(layout = "graphopt", mass = 100) +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, colour = group, tooltip = repo)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common contributors") +
  theme(panel.background = element_rect(fill = "white"))

@@

```{r fig-graph-user-user}
#| fig-cap: Interactive user–user graph. Users are connected by contributions by common repositories, with darker edges indicating more common repositories. Node colour indicates cluster membership according to the edge–betweenness method [girvan2002]. Excludes isolate nodes.
# TODO: is there a khroma scale that can accomodate this many groups?
g_user_user <- oarch_graph_user_user |>
  filter(degree > 170) |>
  ggraph(layout = "graphopt", mass = 100) +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, tooltip = user)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common repositories") +
  theme(panel.background = element_rect(fill = "white"))

girafe(ggobj = g_user_user)
```

<!-- Is there a network? How big is it? -->
Our data shows that there is a significant network of archaeologists collaborating on GitHub.
`r percent(graph_n_repos / graph_n_repos_total)` of repositories and `r percent(graph_n_users / graph_n_users_total)` of users in our dataset are connected to at least one other repository or user.
Of these, `r percent(max(subgraphs_n_repos$n) / graph_n_repos)` of repositories and `r percent(max(subgraphs_n_users$n) / graph_n_users)` of users belong to a single connected subgraph (@fig-graph-repo-repo).

<!-- Clustering within the network -->
To explore the topography of this network, we applied the edge–betweenness community detection method [@girvan2002] to delimit `r nrow(filter(clusters_n_repos, n > 1))` distinct clusters in the repository–repository network.
While many of these clusters are interconnected, some discrete components containing between 2-20 repositories appear as distinct from a primary core.
The core cluster is characterized by repositories whose contributors commit to projects other than their own, and it includes a smorgasbord of projects whose contributors share varied interests.
This indicates the formation of a central software development community rather than a focus on any specific topic of work.
The other peripheral clusters situated around the core largely correspond with the work of single individuals, and sometimes also their close colleagues.
Many of the projects represented in the core emerge from underlying professional partnerships, namely research labs (e.g. ISAA-Kiel) and special interest groups (SSLA).
Peripheral clusters that are connected to core clusters by only a few relationships represent the sole (or perhaps initial) integration of lone developers into a broader community.

@@

The highest mean centrality is seen in repositories that contain lists/datasets, standalone software, or packages/libraries, and in repositories based on Python, R or QGIS [@fig-repo-betweenness-category-platform].
Interestingly, these trends depart from the observed popularity of different categories/platform in the open-archaeo dataset as a whole (see @sec-open-archaeology): standalone software is more central than packages/libraries, even though there are more of the latter by a significant margin.
Similarly, Python packages tend to be more central than R packages, despite being the minority language.
<!-- @ZB: I think this is interesting, but I couldn't really think of a general observation to make here. Maybe you can? -->
<!-- Possibly that R packages are easier to spin up solo, whereas standalone tends to have more financial/institutional backing, and thus greater potential to support teams of developers -->

@@

<!-- Predictors of user centrality -->
Centrality to the user–user graph is weakly predicted by a user's overall rate of activity, as measured by their total number of contributions [@fig-user-betweenness-contribs]. 
We did not collect demographic data on users that appear in our dataset, but based on our own knowledge of the community we can observe that those highly central to the network tend to be employed in (junior) academic positions, or in a few cases in cultural heritage authorities, rather than specifically as research software engineers.
Such positions tend to not to actively reward or encourage software development, at least not on a par with more traditional academic outputs, and are increasingly precarious.
This obviously poses a serious risk to the sustainability and growth of open source software in archaeology: if the people who occupy central positions in the network cease to be active, then it is likely that the overall network would fragment.
Assessing and mitigating this risk should be a high priority for future research in this area.
